Amplicon sequence variants artificially split bacterial genomes into separate clusters

### 2021-02-23 README.Rmd
Use Rcpp to speed up all v all comparison
Deploy generation of confusion matrix data for 100 iterations
Update rrnDB database and NCBI taxonomy

---
title: "Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs"
author: "Pat Schloss"
date: "9/11/2020"
output:
  github_document:
    html_preview: false
---

Developed over a series of *Code Club* episodes led by Pat Schloss to answer an important question in microbiology and develop comfort using tools to develop reproducible research practices.

## Questions
* Within a genome, how many distinct sequences of the 16S rRNA gene are present relative to the number of copies per genome? How far apart are these sequences from each other? How does this scale from a genome to kingdoms?
* Within a taxa (any level), how many ESVs or ASVs from that taxa are shared with sister taxa? How does this change with taxonomic level? Variable region?

```{r library_calls, message=FALSE, echo=FALSE}
library(tidyverse)
library(data.table)
library(rmarkdown)
```

### Dependencies:
* [mothur v.1.44.2](https://github.com/mothur/mothur/tree/v.1.44.2) - `code/install_mothur.sh` installs mothur
* `wget`
* `r R.version.string`
  * `tidyverse` (v. `r packageVersion("tidyverse")`)
  * `data.table` (v. `r packageVersion("data.table")`)
  * `rmarkdown` (v. `r packageVersion("rmarkdown")`)

>>>>

---
title: "Amplicon sequence variants artificially split bacterial genomes into separate clusters"
author: "Pat Schloss"
date: "2/22/2021"
output:
  github_document:
    html_preview: false
---

```{r, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(here))

opts_chunk$set(echo = FALSE)

inline_hook <- function(x){

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps){
			# treat as integer
			formatted <- format(x, digits=0, big.mark = ",")

		} else {
			# treat as floating point number
			formatted <- format(x, digits=2, nsmall=2, big.mark = ",")
		}
	} else {
		formatted <-x
	}

	paste0("**", formatted, "**")

}

knit_hooks$set(inline=inline_hook)

format_pct <- function(x, my_digits=1){

	format(x, digits=my_digits, nsmall=my_digits)

}
```

```{r}
## Paragraph 1
metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character()))

easv <- read_tsv(here("data/processed/rrnDB.easv.count_tibble"),
								col_types = cols(.default = col_character(),
																 count = col_integer()))

metadata_easv <- inner_join(metadata, easv, by=c("genome_id" = "genome")) %>%
	mutate(threshold = recode(threshold, "esv" = "0.000"),
				 threshold = as.numeric(threshold))

rrn_copy_number <- metadata_easv %>%
	filter(region == "v19", threshold == 0) %>%
	group_by(species, genome_id) %>%
	summarize(n_rrn = sum(count), n_asv = n(), .groups="drop") %>%
	group_by(species) %>%
	summarize(min_rrn = min(n_rrn),
            l_quartile_rrn = quantile(n_rrn, prob=0.25),
            median_rrn = median(n_rrn),
            u_quartile_rrn = quantile(n_rrn, prob=0.75),
            max_rrn = max(n_rrn),
            min_asv = min(n_asv),
            l_quartile_asv = quantile(n_asv, prob=0.25),
            median_asv = median(n_asv),
            u_quartile_asv = quantile(n_asv, prob=0.75),
            max_asv = max(n_asv),
            N=n(),
            .groups="drop")

single_copy <- rrn_copy_number %>% filter(median_rrn == 1) %>% top_n(n=1, N)

stopifnot(single_copy$min_rrn == single_copy$max_rrn)

max_copy <- rrn_copy_number %>% top_n(n=1, median_rrn)

single_copy_N <- single_copy %>% pull(N)
#max_copy_N <- max_copy %>% pull(N)

ecoli_copy <- rrn_copy_number %>% filter(species == "Escherichia coli")

rates <- metadata_easv %>%
	filter(threshold == 0) %>%
	group_by(region, genome_id) %>%
	summarize(n_rrns = sum(count), n_asvs = n_distinct(easv), .groups="drop") %>%
	group_by(region) %>%
	summarize(rate = sum(n_asvs)/sum(n_rrns), .groups="drop")

rate_v19 <- rates %>% filter(region == "v19") %>% pull(rate)
rate_v4 <- rates %>% filter(region == "v4") %>% pull(rate)
rate_v34 <- rates %>% filter(region == "v34") %>% pull(rate)
rate_v45 <- rates %>% filter(region == "v45") %>% pull(rate)

asvs_per_species <- metadata_easv %>%
	filter(threshold == 0 & region == "v19") %>%
	group_by(species) %>%
	summarize(n_asvs = n_distinct(easv), .groups="drop")

#below are total number of ASVs across all genomes (above was within a genome)
single_nasvs <- asvs_per_species %>% filter(species == single_copy$species) %>% pull(n_asvs)
ecoli_nasvs <- asvs_per_species %>% filter(species == "Escherichia coli") %>% pull(n_asvs)

## Paragraph 2
thresholds <- read_tsv(here("data/processed/thresholds_for_single_otu.tsv"),
											 col_types=cols(region = col_character(),
											 							 .default = col_double())) %>%
								filter(n_genomes > 100)

threshold_v19 <- thresholds %>% filter(n_rrns == 7 & region == "v19") %>% pull(threshold) * 100
threshold_v4 <- thresholds %>% filter(n_rrns == 7 & region == "v4") %>% pull(threshold) * 100
threshold_v34 <- thresholds %>% filter(n_rrns == 7 & region == "v34") %>% pull(threshold) * 100
threshold_v45 <- thresholds %>% filter(n_rrns == 7 & region == "v45") %>% pull(threshold) * 100

three_v19 <- thresholds %>% filter(threshold <= 0.03 & region == "v19") %>% tail(n=1) %>% pull(n_rrns)
three_v4 <- thresholds %>% filter(threshold <= 0.03 & region == "v4") %>% tail(n=1) %>% pull(n_rrns)
three_v34 <- thresholds %>% filter(threshold <= 0.03 & region == "v34") %>% tail(n=1) %>% pull(n_rrns)
three_v45 <- thresholds %>% filter(threshold <= 0.03 & region == "v45") %>% tail(n=1) %>% pull(n_rrns)

## Paragraph 3
lump_split <- read_tsv(here("data/processed/lumped_split_rate.tsv"),
											 col_types = cols(region = col_character(),
											 								 .default = col_double()))

lump_v19 <- lump_split %>% filter(threshold ==0 &region == "v19") %>% pull(lump_rate) * 100
lump_v4 <- lump_split %>% filter(threshold ==0 &region == "v4") %>% pull(lump_rate) * 100
lump_v34 <- lump_split %>% filter(threshold ==0 &region == "v34") %>% pull(lump_rate) * 100
lump_v45 <- lump_split %>% filter(threshold ==0 &region == "v45") %>% pull(lump_rate) * 100

lump3_v19 <- lump_split %>% filter(threshold ==0.03 & region == "v19") %>% pull(lump_rate) * 100
lump3_v4 <- lump_split %>% filter(threshold ==0.03 & region == "v4") %>% pull(lump_rate) * 100
lump3_v34 <- lump_split %>% filter(threshold ==0.03 & region == "v34") %>% pull(lump_rate) * 100
lump3_v45 <- lump_split %>% filter(threshold ==0.03 & region == "v45") %>% pull(lump_rate) * 100

lump_split_iqr <- lump_split %>% pivot_longer(cols=c(split_iqr, lump_iqr), names_to="test", values_to="iqr") %>% summarize(max(iqr)) %>% pull

```

## Abstract

Amplicon sequencing variants (ASVs) have been proposed as an alternative to operational taxonomic units (OTUs) for analyzing microbial communities. ASVs have grown in popularity, in part, because of a desire to reflect a more refined level of taxonomy since they do not cluster sequences based on a distance-based threshold. However, ASVs and the use of overly narrow thresholds to identify OTUs increase the risk of splitting a single genome into separate clusters. To assess this risk, I analyzed the intragenomic variation of 16S rRNA genes from the bacterial genomes represented in a *rrn* copy number database, which contained `r nrow(metadata)` genomes from `r n_distinct(metadata$species)` species. As the number of copies of the 16S rRNA gene increased in a genome, the number of ASVs also increased. There was an average of `r rate_v19` ASVs per copy of the 16S rRNA gene for full length 16S rRNA genes. It was necessary to use a distance threshold of `r format_pct(threshold_v19)`% to cluster full length ASVs from the same genome into a single OTU with 95% confidence for genomes with 7 copies of the 16S rRNA, such as *E. coli*. This research highlights the risk of splitting a single bacterial genome into separate clusters when ASVs are used to analyze 16S rRNA gene sequence data. Although there is also a risk of clustering ASVs from different species into the same OTU when using broad distance thresholds, those risks are of less concern than artificially splitting a genome into separate ASVs and OTUs.

## Importance

16S rRNA gene sequencing has engendered significant interest in studying microbial communities. There has been a tension between trying to classify 16S rRNA gene sequences to increasingly lower taxonomic levels and the reality that those levels were defined using more sequence and physiological information than is available from a fragment of the 16S rRNA gene. Furthermore, naming of bacterial taxa reflects the biases of those who name them. One motivation for the recent push to adopt ASVs in place of OTUs in microbial community analyses is to allow researchers to perform their analyes at the finest possible level that reflects species-level taxonomy. The current research is significant because it quantifies the risk of artificially splitting bacterial genomes into separate clusters. Far from providing a better represenation of bacterial taxonomy and biology, the ASV approach can lead to conflicting inferences about the ecology of different ASVs from the same genome.

```{r library_calls, message=FALSE, echo=FALSE}
library(tidyverse)
library(data.table)
library(rmarkdown)
library(Rcpp)
library(furrr)
```

### Dependencies:
* [mothur v.1.44.2](https://github.com/mothur/mothur/tree/v.1.44.2) - `code/install_mothur.sh` installs mothur
* `wget`
* `r R.version.string`
  * `tidyverse` (v. `r packageVersion("tidyverse")`)
  * `Rcpp` (v. `r packageVersion("Rcpp")`)
  * `furrr` (v. `r packageVersion("furrr")`)
  * `data.table` (v. `r packageVersion("data.table")`)
  * `rmarkdown` (v. `r packageVersion("rmarkdown")`)

### 2021-02-23 submission/manuscript.Rmd
Update rrnDB database and NCBI taxonomy

format_pct <- function(x, my_digits=1){

@@

**Materials and Methods. (i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.6, released November 8, 2019) [@Stoddard2014]. At the time of submission, this was the most current version of the database. The *rrn*DB obtained the curated 16S rRNA gene sequences from the KEGG database, which ultimately obtained them from NCBI's non-redundant RefSeq database. The *rrn*DB provided downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy. For some genomes this resulted in multiple classifications since a genome's 16S rRNA gene sequences were not identical. Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome. Because strain names were not consistently given to genomes across bacterial species, I disregarded the strain level designations.

@@

**(iii) Clustering sequences into OTUs.** Pairwise distances between sequences were calculated using the dist.seqs command from mothur. The OptiClust algorithm as implemented in mothur was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017]. Distance thresholds between 0.5 and 10.0% in 0.5 percentage point increments were used to assign sequences to OTUs.

>>>>

format_pct <- function(x, my_digits=2){

@@

**Materials and Methods. (i) Data availability.** The 16S rRNA gene sequences used in this study were obtained from the *rrn*DB (https://rrndb.umms.med.umich.edu; version 5.7, released January 18, 2021) [@Stoddard2014]. At the time of submission, this was the most current version of the database. The *rrn*DB obtained the curated 16S rRNA gene sequences from the KEGG database, which ultimately obtained them from NCBI's non-redundant RefSeq database. The *rrn*DB provided downloadable versions of the sequences with their taxonomy as determined using the naive Bayesian classifier trained on the RDP reference taxonomy. For some genomes this resulted in multiple classifications since a genome's 16S rRNA gene sequences were not identical. Instead, I mapped the RefSeq accession number for each genome in the database to obtain a single taxonomy for each genome. Because strain names were not consistently given to genomes across bacterial species, I disregarded the strain level designations.

@@

**(iii) Clustering sequences into OTUs.** Pairwise distances between sequences were calculated using the dist.seqs command from mothur. The OptiClust algorithm as implemented in mothur was used to assign 16S rRNA gene sequences to OTUs [@Westcott2017]. Distance thresholds between 0.25 and 10.0% in 0.25 percentage point increments were used to assign sequences to OTUs.

### 2021-02-23 submission/manuscript.tex
Update rrnDB database and NCBI taxonomy

Amplicon sequencing variants (ASVs) have been proposed as an alternative
to operational taxonomic units (OTUs) for analyzing microbial
communities. ASVs have grown in popularity, in part, because of a desire
to reflect a more refined level of taxonomy since they do not cluster
sequences based on a distance-based threshold. However, ASVs and the use
of overly narrow thresholds to identify OTUs increase the risk of
splitting a single genome into separate clusters. To assess this risk, I
analyzed the intragenomic variation of 16S rRNA genes from the bacterial
genomes represented in a \emph{rrn} copy number database, which
contained \textbf{15,614} genomes from \textbf{4,774} species. As the
number of copies of the 16S rRNA gene increased in a genome, the number
of ASVs also increased. There was an average of \textbf{0.60} ASVs per
copy of the 16S rRNA gene for full length 16S rRNA genes. It was
necessary to use a distance threshold of \textbf{5.5}\% to cluster full
length ASVs from the same genome into a single OTU with 95\% confidence
for genomes with 7 copies of the 16S rRNA, such as \emph{E. coli}. This
research highlights the risk of splitting a single bacterial genome into
separate clusters when ASVs are used to analyze 16S rRNA gene sequence
data. Although there is also a risk of clustering ASVs from different
species into the same OTU when using broad distance thresholds, those
risks are of less concern than artificially splitting a genome into
separate ASVs and OTUs.

@@

To investigate the variation in the number of copies of the 16S rRNA
gene per genome and the intragenomic variation among copies of the 16S
rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
database (\emph{rrn}DB)(14). Among the \textbf{4,774} species
represented in the \emph{rrn}DB there were \textbf{15,614} genomes. The
median \emph{rrn} copy number per species ranged between \textbf{1}
(e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}). As \textbf{rrn} copy number
for a genome increased, the number of variants of the 16S rRNA gene in
each genome also increased. On average, there were \textbf{0.60}
variants per copy of the full length 16S rRNA gene and an average of
\textbf{0.33}, \textbf{0.26}, and \textbf{0.27} variants when
considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
Although a species tended to have a consistent number of 16S rRNA gene
copies per genome, the number of total variants increased with the
number of genomes that were sampled (\textbf{Figure 1}). For example,
the \textbf{180} genome accessions of \textbf{\emph{Mycobacterium
tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
per genome. However, across those accessions, there were \textbf{11}
versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
copies of the 16S rRNA gene with a median of \textbf{5} distinct full
length ASVs per genome (intraquartile range between \textbf{4} and
\textbf{6}). Across the \textbf{958} \emph{E. coli} genomes in the
\emph{rrn}DB, there were \textbf{1,013} versions of the gene. These
observations highlight the risk of selecting a threshold for defining
clusters that is too narrow because it is possible to split a single
genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to
cluster 16S rRNA gene sequences together based on their distances
between each other. Therefore, I assessed the impact of the distance
threshold used to define clusters of 16S rRNA genes on the propensity to
split a genome into separate clusters. To control for uneven
representation of genomes across species, I randomly selected one genome
from each species for the following analyses and repeated each
randomization 100 times. I observed that as the \emph{rrn} copy number
increased, the distance threshold required to reduce the ASVs in each
genome to a single OTU increased (Figure 1). Among species with 7 copies
of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
\textbf{5.5}\% was required to reduce full length ASVs into a single OTU
for 95\% of the species (Figure 2). Similarly, thresholds of
\textbf{4.0}, \textbf{2.5}, and \textbf{3.5}\% were required for the
V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
threshold was used, then ASVs from genomes containing fewer than
\textbf{5}, \textbf{6}, \textbf{8}, and \textbf{6} copies of the
\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
Consequently, these results demonstrate that broad thresholds must be
used to avoid splitting different operons from the same genome into
separate clusters. At broad thresholds, 16S rRNA gene sequences from
multiple species could be clustered into the same ASV or OTU
(\textbf{Figure 2}). Using ASVs, \textbf{3.6}\% of the ASVs contained
sequences from multiple species when considering full length sequences
and \textbf{10.2}, \textbf{14.9}, and \textbf{12.0}\% when considering
the V3-V4, V4, and V4-V5 regions, respectively. At the commonly used 3\%
threshold for defining OTUs, \textbf{25.2}\% of the OTUs contained 16S
rRNA gene sequences from multiple species when considering full length
sequences and \textbf{29.4}, \textbf{33.0}, and \textbf{32.2}\% when
considering the V3-V4, V4, and V4-V5 regions, respectively. Considering
that species designations are unevenly applied and reflect multiple
human-imposed biases, the risk of splitting a genome into multiple OTUs
is more problematic than clustering species together. Therefore, larger
thresholds are advisable.

@@

\textbf{Materials and Methods. (i) Data availability.} The 16S rRNA gene
sequences used in this study were obtained from the \emph{rrn}DB
(\url{https://rrndb.umms.med.umich.edu}; version 5.6, released November
8, 2019) (14). At the time of submission, this was the most current
version of the database. The \emph{rrn}DB obtained the curated 16S rRNA
gene sequences from the KEGG database, which ultimately obtained them
from NCBI's non-redundant RefSeq database. The \emph{rrn}DB provided
downloadable versions of the sequences with their taxonomy as determined
using the naive Bayesian classifier trained on the RDP reference
taxonomy. For some genomes this resulted in multiple classifications
since a genome's 16S rRNA gene sequences were not identical. Instead, I
mapped the RefSeq accession number for each genome in the database to
obtain a single taxonomy for each genome. Because strain names were not
consistently given to genomes across bacterial species, I disregarded
the strain level designations.

@@

\textbf{(iii) Clustering sequences into OTUs.} Pairwise distances
between sequences were calculated using the dist.seqs command from
mothur. The OptiClust algorithm as implemented in mothur was used to
assign 16S rRNA gene sequences to OTUs (27). Distance thresholds between
0.5 and 10.0\% in 0.5 percentage point increments were used to assign
sequences to OTUs.

\textbf{(iv) Controlling for uneven sampling of genomes by species.}
Because of the uneven distribution of genome sequences across species,
for the analysis of splitting genomes and clustering ASVs from different
species, I randomly selected one genome from each species. The random
selection was repeated 100 times. Analyses based on this randomization
reported the median of the 100 randomizations. The intraquartile range
between randomizations was less than \textbf{0.0029}. Because the range
was so small, the confidence intervals were more narrow than the
thickness of the lines in Figures 1 and 2 and were not included.

>>>>

Amplicon sequencing variants (ASVs) have been proposed as an alternative
to operational taxonomic units (OTUs) for analyzing microbial
communities. ASVs have grown in popularity, in part, because of a desire
to reflect a more refined level of taxonomy since they do not cluster
sequences based on a distance-based threshold. However, ASVs and the use
of overly narrow thresholds to identify OTUs increase the risk of
splitting a single genome into separate clusters. To assess this risk, I
analyzed the intragenomic variation of 16S rRNA genes from the bacterial
genomes represented in a \emph{rrn} copy number database, which
contained \textbf{20,427} genomes from \textbf{5,972} species. As the
number of copies of the 16S rRNA gene increased in a genome, the number
of ASVs also increased. There was an average of \textbf{0.58} ASVs per
copy of the 16S rRNA gene for full length 16S rRNA genes. It was
necessary to use a distance threshold of \textbf{5.25}\% to cluster full
length ASVs from the same genome into a single OTU with 95\% confidence
for genomes with 7 copies of the 16S rRNA, such as \emph{E. coli}. This
research highlights the risk of splitting a single bacterial genome into
separate clusters when ASVs are used to analyze 16S rRNA gene sequence
data. Although there is also a risk of clustering ASVs from different
species into the same OTU when using broad distance thresholds, those
risks are of less concern than artificially splitting a genome into
separate ASVs and OTUs.

@@

To investigate the variation in the number of copies of the 16S rRNA
gene per genome and the intragenomic variation among copies of the 16S
rRNA gene, I obtained 16S rRNA sequences from the \emph{rrn} copy number
database (\emph{rrn}DB)(14). Among the \textbf{5,972} species
represented in the \emph{rrn}DB there were \textbf{20,427} genomes. The
median \emph{rrn} copy number per species ranged between \textbf{1}
(e.g., \textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}). As \textbf{rrn} copy number
for a genome increased, the number of variants of the 16S rRNA gene in
each genome also increased. On average, there were \textbf{0.58}
variants per copy of the full length 16S rRNA gene and an average of
\textbf{0.32}, \textbf{0.25}, and \textbf{0.27} variants when
considering the V3-V4, V4, and V4-V5 regions of the gene, respectively.
Although a species tended to have a consistent number of 16S rRNA gene
copies per genome, the number of total variants increased with the
number of genomes that were sampled (\textbf{Figure 1}). For example,
the \textbf{271} genome accessions of \textbf{\emph{Mycobacterium
tuberculosis}} in the \emph{rrn}DB each had \textbf{1} copy of the gene
per genome. However, across those accessions, there were \textbf{17}
versions of the gene. An \emph{E. coli} genome typically had \textbf{7}
copies of the 16S rRNA gene with a median of \textbf{5} distinct full
length ASVs per genome (intraquartile range between \textbf{3} and
\textbf{6}). Across the \textbf{1,390} \emph{E. coli} genomes in the
\emph{rrn}DB, there were \textbf{1,402} versions of the gene. These
observations highlight the risk of selecting a threshold for defining
clusters that is too narrow because it is possible to split a single
genome into multiple clusters.

A method to avoid splitting a single genome into multiple clusters is to
cluster 16S rRNA gene sequences together based on their distances
between each other. Therefore, I assessed the impact of the distance
threshold used to define clusters of 16S rRNA genes on the propensity to
split a genome into separate clusters. To control for uneven
representation of genomes across species, I randomly selected one genome
from each species for the following analyses and repeated each
randomization 100 times. I observed that as the \emph{rrn} copy number
increased, the distance threshold required to reduce the ASVs in each
genome to a single OTU increased (Figure 1). Among species with 7 copies
of the \emph{rrn} operon (e.g., \emph{E. coli}), a distance threshold of
\textbf{5.25}\% was required to reduce full length ASVs into a single
OTU for 95\% of the species (Figure 2). Similarly, thresholds of
\textbf{5.25}, \textbf{2.50}, and \textbf{3.75}\% were required for the
V3-V4, V4, and V4-V5 regions, respectively. But, if a 3\% distance
threshold was used, then ASVs from genomes containing fewer than
\textbf{10}, \textbf{10}, \textbf{8}, and \textbf{10} copies of the
\emph{rrn} operon would reliably be clustered into a single OTU for ASVs
from the V1-V9, V3-V4, V4, and V4-V5 regions, respectively (Figure 2).
Consequently, these results demonstrate that broad thresholds must be
used to avoid splitting different operons from the same genome into
separate clusters. At broad thresholds, 16S rRNA gene sequences from
multiple species could be clustered into the same ASV or OTU
(\textbf{Figure 2}). Using ASVs, \textbf{4.11}\% of the ASVs contained
sequences from multiple species when considering full length sequences
and \textbf{10.87}, \textbf{16.20}, and \textbf{13.06}\% when
considering the V3-V4, V4, and V4-V5 regions, respectively. At the
commonly used 3\% threshold for defining OTUs, \textbf{27.41}\% of the
OTUs contained 16S rRNA gene sequences from multiple species when
considering full length sequences and \textbf{31.71}, \textbf{34.31},
and \textbf{34.78}\% when considering the V3-V4, V4, and V4-V5 regions,
respectively. Considering that species designations are unevenly applied
and reflect multiple human-imposed biases, the risk of splitting a
genome into multiple OTUs is more problematic than clustering species
together. Therefore, larger thresholds are advisable.

@@

\textbf{Materials and Methods. (i) Data availability.} The 16S rRNA gene
sequences used in this study were obtained from the \emph{rrn}DB
(\url{https://rrndb.umms.med.umich.edu}; version 5.7, released January
18, 2021) (14). At the time of submission, this was the most current
version of the database. The \emph{rrn}DB obtained the curated 16S rRNA
gene sequences from the KEGG database, which ultimately obtained them
from NCBI's non-redundant RefSeq database. The \emph{rrn}DB provided
downloadable versions of the sequences with their taxonomy as determined
using the naive Bayesian classifier trained on the RDP reference
taxonomy. For some genomes this resulted in multiple classifications
since a genome's 16S rRNA gene sequences were not identical. Instead, I
mapped the RefSeq accession number for each genome in the database to
obtain a single taxonomy for each genome. Because strain names were not
consistently given to genomes across bacterial species, I disregarded
the strain level designations.

@@

\textbf{(iii) Clustering sequences into OTUs.} Pairwise distances
between sequences were calculated using the dist.seqs command from
mothur. The OptiClust algorithm as implemented in mothur was used to
assign 16S rRNA gene sequences to OTUs (27). Distance thresholds between
0.25 and 10.0\% in 0.25 percentage point increments were used to assign
sequences to OTUs.

\textbf{(iv) Controlling for uneven sampling of genomes by species.}
Because of the uneven distribution of genome sequences across species,
for the analysis of splitting genomes and clustering ASVs from different
species, I randomly selected one genome from each species. The random
selection was repeated 100 times. Analyses based on this randomization
reported the median of the 100 randomizations. The intraquartile range
between randomizations was less than \textbf{0.0024}. Because the range
was so small, the confidence intervals were more narrow than the
thickness of the lines in Figures 1 and 2 and were not included.
